---
title: "Introduction"
author: "Mirjam Laager"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true

vignette: >
  %\VignetteIndexEntry{Introduction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  
bibliography: bibliography.bib
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

In this tutorial we will be analysing the dataset `patient_data_simulated` to 
demonstrate the main functionalities of the package *mrsamcmc*. 


# Example Dataset
The dataset `patient_data_simulated` is distributed with this package and contains
simulated admission and discharge dates, test results and antibiotic use of 
270 patients over 90 days. 
```{r}
library(mrsamcmc)
example_dataset <- patient_data_simulated
```

Data to be analysed with *mrsamcmc* should be in
the same format as the example dataset we use here. So let's look at the example
dataset in more detail. It is a list with four elements: `$patients`, 
`$test_results_positive`, `$test_results_negative` and `$antibiotics`.

## patients
```{r, look at example dataset}
head(example_dataset$patients)
```
In `$patients` each row corresponds to a patient and the columns are the admission
and discharge dates of the patient (in days) and the day of the first positive test 
for those patients who were tested positive. If a patient was never tested positive, 
their day of first positive test is set to 20000. The dataset also
contains a column for the types of the patients. This column is ignored in the
simplest form of the model `run_mcmc()`, which does not infer sources of infection.

## test results
In `$test_results_positive` and `$test_results_negative` each row corresponds to a 
patient and the entries are the day of a positive or negative test for that patient.
We can get the total number of positive and negative tests in the example 
dataset as follows:
```{r, get positive and negative tests in example dataset}
length(which(example_dataset$test_results_positive != 0))
length(which(example_dataset$test_results_negative != 0))
```

## antibiotics
In `$antibiotics` each row corresponds to a patient and the entries are the days 
on which antibiotics were administered for that patient. Patient 4 for example
was admitted on day `r example_dataset$patients[4,]$admission` and
discharged on day `r example_dataset$patients[4,]$discharge` and was
on antibiotics on the following days:
```{r, get antibiotics in example dataset}
example_dataset$antibiotics[4,example_dataset$antibiotics[4,] != 0]
```

# Model
The model implemented in the package *mrsamcmc* very much builds on previous
methodology as described in @doi:10.1093/biostatistics/kxl017 and @pmid27042253.
The data augmentation approach used in these publications introduces
additional parameters for the unknown colonisation time for each patient, which
makes it possible to split the overall likelihood into the following product:
$$ \Pi(\Omega, W| \rho, \varphi, \beta) = \Pi(\Omega|W,\rho) \cdot \Pi(W | \rho, \beta),  $$
where $\Omega$ denotes the test results, W is the colonisation status of the patients and colonisation times, $\rho$ is the test sensitivity, $\varphi$ is the probability of a patient being 
positive on admission and $\beta$ is a transmission parameter. The first factor of the product describes the observation process, which can be modelled as a binomial distribution. Assuming perfect specificity, the likelihood of the test results given the test sensitivity and the true statuses is therefore given by  

$$ \Pi\left(\Omega\middle|W,\rho\right)=\rho^{TP}\cdot\left(1-\rho\right)^{FN} $$

where TP and FN denote the number of true positive and false negative tests. The second factor of the overall likelihood captures the transmission process. Following  @pmid27042253 we assume that patients who acquire a colonisation stay colonised until discharge. The patient statuses, W, are therefore fully described by the colonisation times, $t^c$, which are set to infinity for patients who do not get colonised during their ward stay. By discretising time into days, the likelihood of the colonisation times given the transmission parameters can be modelled as

$$ \Pi\left(t^c\middle|\varphi,\beta\right)=\varphi^{N_p}\left(1-\varphi\right)^{N-N_p}\prod_{k:t_k^c=\infty}\left[\prod_{j=t_k^a}^{t_k^d}\left(1-p_{kj}\right)\right]\prod_{k:t_k^c\neq\infty}\left[p_{kt_k^c}\prod_{j=t_k^a}^{t_k^c-1}\left(1-p_{kj}\right)\right] $$


where $N$ is the total number of patients, $N_p$ is the number of patients who are admitted already colonised and $t_k^a$, $t_k^c$ and $t_k^d$ denote the day of admission, colonisation and discharge of patient $k$. $p_{kj}$ is the probability of patient $k$ becoming colonised on day $j$ which is given by

$$ p_{kj}= 1 - \exp(- \beta C(j)) $$

where $C(j)$ is the number of colonised patients on day $j$.




## Accounting for the effect of antibiotics

We expand this model to account for the effect of antibiotics on detection and transmission. 
We assume that antibiotics can affect the test sensitivity, the probability of acquisition and the probability of onward transmission. To capture the effect of antibiotics on test sensitivity we assume that in the presence of antibiotics the baseline test sensitivity changes by a factor \delta and therefore model the likelihood of the test results as

$$ \Pi\left(\Omega\middle|W,\rho,\delta,\Sigma\right)=\rho^{TP_{nabx}}\cdot\left(1-\rho\right)^{FN_{nabx}}\cdot\left(\delta\rho\right)^{TP_{abx}}\cdot\left(1-\delta\rho\right)^{FN_{abx}} $$

where $TP_{nabx}$, ${FN}_{nabx}$, $TP_{abx}$ and ${FN}_{abx}$ denote the total number of true positive and false negative tests conducted while a patient was on or off antibiotics and $\Sigma$ is a matrix with entries $\sigma_{ij}$ equal to one if patient i is on antibiotics on day j and zero otherwise. We model the effect of antibiotics on the transmission dynamics by introducing a parameter $\alpha$ for the effect on acquisition and a parameter $\tau$ for the effect on onward transmission and modify the daily probability of acquisition as follows

$$ p_{kj}=1-\ exp(\ -\ \alpha^{1_{\sigma_{kj}=1}}(\beta C_{nabx}\left(j\right)+\tau\beta C_{abx}\left(j\right))) $$

where $C_{nabx}(j)$ and $C_{abx}(j)$ denote the number of colonised patients on day j who are on or off antibiotics and $1_{\sigma_{kj}=1}$ is an indicator function which equals to 1 if patient $k$ is on antibiotics on day $j$ and zero otherwise. 


# Inference with customised settings
In this section we will run the MCMC algorithm with the example dataset and
show how to customise the settings of the MCMC and use custom prior distributions.
The function `run_mcmc()` has three arguments, `data`, `priors` and `configuration`.
If none of these arguments are specified, the function runs with default priors and
configuration. Here, we will use some custom settings.

## Customise configuration
We generate a custom configuration using `generate_configuration()`. This function
returns the default configuration, which can be inspected as follows:
```{r, get default configuration}
my_configuration <- generate_configuration()
my_configuration$n_iter
```
For the purpouse of this tutorial, we run a short chain with 3000 iterations only. We
also infer no effects of antibiotics for now and modify the output storage such
that we only store aggreated patient data only .

```{r, customise configuration using fuction}
my_configuration <- generate_configuration(n_iter = 3000, store_params = "STORE_ALL", 
store_statuses = "AGGREGATED", infer_covariate_effects = "NONE")
```

Alternatively, we can get the default configuration and modify entries individually
```{r, customise configuration alternative}
new_configuration <- generate_configuration()
new_configuration$n_iter <- 300
```


## Customise priors
We generate custom priors using `generate_priors()`. This function returns
the default priors, which can be inspected as follows:
```{r, get default priors}
my_priors <- generate_priors()
my_priors$functions$f_beta
my_priors$parameters$p_beta
```
The default prior for the transmission rate is a Cauchy distribution with
parameters 0 and 4. Let's change that to a Uniform prior:

```{r, customise priors modifying default}
my_priors$functions$f_beta <- dunif
my_priors$parameters$p_beta <- c(0,1)
```
Alternatively, we can generate the custom priors with the `generate_priors()` 
function directly:

```{r, customise priors alternative}
my_priors <- generate_priors(f_beta = dunif, p_beta = c(0,1) )

```

## Run MCMC
Now, we can run the MCMC using our customised configuration and priors:
```{r, run mcmc}
output <- run_mcmc(data = example_dataset, priors = my_priors, 
configuration = my_configuration)
```



# Analysing the output
We use the package *coda* to take a look at the results. Because the MCMC augments
the true colonisation times, we can now estimate, what proportion of the colnised
patients were admitted already colonised and what proportion acquired the colonisation
during their stay:
```{r, analyse output}
library(coda)
library(ggplot2)
print(HPDinterval(as.mcmc(output$statuses)))
```
We can also plot the chains and densities of individual parameters. Since we only ran
a small number of iterations for the purpose of this tutorial, the chain clearly has not
yet converged. Consider running more iterations (for example 30'000) and running multiple
chains for better results.
```{r, plot output, fig.width = 7}
full_chain_beta <- output$parameters$beta
chain_without_burnin_beta <- full_chain_beta[-(1:my_configuration$n_burnin)]
plot(as.mcmc(chain_without_burnin_beta))
```

For further output analysis including posterior predicitive checks and 
forward simulations, consider the vignette *simulations*.


# References



