---
title: "Introduction"
author: "Mirjam Laager"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true

vignette: >
  %\VignetteIndexEntry{Introduction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

In this tutorial we will be analysing the dataset `patient_data_simulated` to 
demonstrate the main functionalities of the package *mrsamcmc*. 


# Example Dataset
The dataset `patient_data_simulated` is distributed with this package and contains
simulated admission and discharge dates, test results and antibiotic use of 
450 patients over 90 days. Data to be analysed with *mrsamcmc* should be in
the same format as the example dataset we use here. So let's look at the example
dataset in more detail. It is a list with four elements: `$patients`

Here: explain the data format


```{r}
library(mrsamcmc)
example_dataset <- patient_data_simulated
head(example_dataset$patients)
```

# Brief overview over the MCMC algorithm
Here: describe what it does. The likelihood functions and cite the papers.



# Running the MCMC with customised settings
Here: show how to generate config and priors

# Analysing the output
Here: do some plots of chains (just show that can use coda)
Also: do some convergence analysis (just show that can use the diagnostics)
Also: run posterior predictive checks











